### 2015-12-05 mcmc.tex
words about initialization

% style notes
% -----------
% - These documents are ``Chapters'', no?
% - [LaTeX] newline after every full stop for proper git diffing
% - [LaTeX] eqnarray not equation
% - pdf not PDF
% - is it ``a MCMC sampling'' or ``an MCMC sampling''?
% - \note{} right after word if you are endnoting a phrase,
%   but after the full stop if endnoting the sentence or paragraph.
% - always $\pars$ never $x$; $K$ samples $\pars_k$ and $\pars$ space is $D$-dimensional
% - should we use the word ``walker'' anywhere, everywhere, or never?

@@

@@

...Why does M--H MCMC work (briefly)? ...Or put this in a note on an earlier paragraph?

@@

Just as most (though not all) MCMC methods require a choice of proposal distribution,
  most (though not all) require a choice about initialization:
The walker needs (or walkers need) to be started somewhere in the parameter space.
In many use cases for MCMC,
  the investigator wants to use it to obtain uncertainty information about
  or propagate uncertainty into parameter estimates.
In these cases, it makes sense to initialize the walker or walkers at sensible parameter estimates,
  found by optimizing a likelihood or a posterior PDF in advance of sampling.
In extremely high dimensions (large numbers of parameters), this can be a bad idea,
  since the optimal parameters are not necessarily near typical posterior samples\note{Make the counterintuitive point.},
  but in low dimensions (few to tens) this is often sensible.

In other use cases,
  the investigator wants to use MCMC to search all of parameter space
  for good models, and also sample.
In this case, it makes sense to be more catholic with initialization...

...but note that MCMC is not the ideal search algorithm for any parameter space.

...Start from different locations
% BJB
% What different locations?
% Drawn from the prior can be useful. Then you can
% look at the chains and see which ones "made it" and
% which ones "got stuck" and delete the latter

...Dealing with multi-modal problems
% Starting from different locations is good
% for DIAGNOSING a problem as multimodal but it's
% terrible for quantifying the modes. Explain why,
% it's a good teaching point.

>>>>

% style notes
% -----------
% - These documents are ``Chapters'', no?
% - [LaTeX] newline after every full stop for proper git diffing
% - [LaTeX] eqnarray not equation
% - pdf not PDF
% - is it ``a MCMC sampling'' or ``an MCMC sampling''?
% - \note{} right after word if you are endnoting a phrase,
%   but after the full stop if endnoting the sentence or paragraph.
% - always $\pars$ never $x$; $K$ samples $\pars_k$ and $\pars$ space is $D$-dimensional
% - should we use the word ``walker'' everywhere and do we?

@@

DWH: ...Why does M--H MCMC work (briefly)?  Here or later?

@@

@@

Just as most (though not all) MCMC methods require a choice of
proposal distribution, most (though not all) require a choice about
initialization:
The walker needs (or walkers need) to be started somewhere in the
parameter space.
In many use cases for MCMC, the investigator wants to use it to obtain
uncertainty information about or propagate uncertainty into parameter
estimates.
In these cases, it makes sense to initialize the walker or walkers at
sensible parameter estimates, found by optimizing a likelihood or a
posterior PDF in advance of sampling.
In extremely high dimensions (large numbers of parameters), this can
be a bad idea, since the optimal parameters are not necessarily near
typical posterior samples\note{When the dimensionality of the
  parameter space gets very large, almost all samples will be
  \emph{very far} from the optimum of the posterior pdf.  One way to
  think about this is to think of a random Gaussian draw in $D$
  dimensions: A typical draw will be $\sqrt{D}$ standard deviations
  Euclidean distance from the maximum of the Gaussian.  That's a long
  way when $D=100$ or $1000$!  Another way to think about it is that
  there are a lot of ways to move \emph{away} from the point you care
  about, but very few ways to move close to it.  The upshot is that
  even with an \emph{enormous} sampling, if you are in large
  dimensions, not a single sample will be very near the optimum.}, but
in low dimensions (few to tens) this is often sensible.

Ideally, you will initialize the walker not at a completely irrelevant
point, nor at the optimum of the posterior pdf, but at a typical or
pretty good place in the posterior pdf.
That would minimize burn-in!
But, in general, being at the optimum or anywhere near a good place is
usually better than a mindless initialization.

If you suspect, or if it is even possible, that your problem is badly
multi-modal, then you will have to start at multiple points in
parameter space and compare the resulting chains.
By ``badly multi-modal'' we mean that there are peaks in the posterior
pdf that are connected by low-enough valleys that it is very unlikely
or takes a long time for a walker to traverse from one peak to the
next.
If this might be an issue, then the best diagnosis is to start many
chains in parallel and check that they lead to identical (or very
similar) posterior inferences (same pdf location and shape and
statistics).

Of course if you find out the worst---if you find out that different
initializations lead to different posterior inferences---you are in
trouble:
There is no trivial way to combine together the samplings you get of
different modes!
You can either wait a \emph{very long time} so that you see a single
walker in a single MCMC run traverse from mode to mode enough times to
make a representative sampling (dozens or hundreds of times, ideally),
or you can use a high-end sampler (like nested sampling) that is
designed for such problems.
In principle there are methods that involve splitting the space into
two spaces, sampling them separately, and then combining the chains
according to their relative evidence afterwards.
That's a research project beyond the scope of this \documentname.

DFM: IS THIS PARAGRAPH USEFUL OR CLEAR?
Of course there are some kinds of multi-modalities that aren't a
problem.
For example, you can have two solutions to a problem that differ only
in the \emph{labeling} of the components---say the Gaussians making up
a mixture of Gaussians---that make up the model.
One solution has the $K$ Gaussian components in one order, and the
other in another order, but they are identical Gaussians in all
other respects.
Because these two solutions make the same predictions for any data, and
only differ in their irrelevant, latent, internal ``naming'' of components,
they are not really different solutions.
This suggests that when you ask whether the outputs of two MCMC chains
are consistent with one another, you should do so in the realm of the
parameters you \emph{care about}, not irrelevant parameters that do
not have impact on any present or future data.

But it is worth remembering that just as MCMC is not a good optimizer
(generically, samples will not lie close to the maximum of the posterior pdf),
it is also not a good \emph{search algorithm}.
There is no sense in which (standard) MCMC methods are efficient at
searching, or engineered to search, all of parameter space.
If you really need to check all of parameter space, you should put
MCMC aside, and do an exhaustive search.\note{You might want to open a
  bottle of wine or two while the code runs.}
